---
title: "Spotify Data Analysis"
author: "James Akemu"
date: "2023-12-29"
output: html_document
---

# Spotify Data Analysis

This dataset contains a comprehensive list of the most famous songs from 1970 to 2023 as listed on Spotify.

## Load the Necessary Packages.

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(knitr)
```

## Import and Clean our Data

```{r}
spotify <- read.csv('spotify-2023.csv') # Import the data
str(spotify)
spotify$artist_count <- as.factor(spotify$artist_count)
spotify$streams <- as.integer(spotify$streams)
spotify$in_deezer_playlists <- as.numeric(spotify$in_deezer_playlists)
spotify$in_shazam_charts <- as.numeric(spotify$in_shazam_charts)
spotify$mode <- as.factor(spotify$mode)
label <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
spotify$released_month <- factor(spotify$released_month, labels = label, ordered = TRUE)
head(spotify)
```

Let's Have a summary of the Data

```{r}
summary(spotify)
```

## Most popular artists and songs

```{r}
by_stream <- spotify %>% group_by(artist.s._name) %>% summarise(sum_streams = sum(streams)) %>% 
  arrange(desc(sum_streams))
head(by_stream)
```

Let us Graph the six most Popular Artists

```{r}
ggplot(head(by_stream), mapping = aes(x = artist.s._name, y = sum_streams)) + 
  geom_col(fill = 'blue') + ggtitle('Most Popular Artists') + 
  labs(x = 'Artist', y = 'Total Streams')
```

Taylor Swift is the most popular artist on Spotify with the most streams.

The most popular song:

```{r}
pop_song <- spotify %>% select(track_name, artist.s._name, released_year, streams) %>% 
  arrange(desc(streams))
head(pop_song)
```

The most popular song is 'take me to church' by Hozier in 2013.

## Time Series Analysis

### Number of songs released on Spotify by year.

```{r}
gg <- spotify %>% group_by(released_year) %>% summarise(count_num_of_song = n())
head(gg)
```

The Year with the most songs released was 2022.

```{r}
gg %>% arrange(desc(count_num_of_song))
```

Let us Graph song releases year on

```{r}
plot(x = gg$released_year, y = gg$count_num_of_song, type = 'l', lwd = 2, xlab = 'year',
ylab = 'No of released songs', main = 'No of Released songs per year', sub = '1930-2023')
```

The number of songs released increased rapidly in 2021 peaking in 2022, and then slowing down in 2023.

Let's analyze the year 2022(the year with the most songs).

```{r}
month <- spotify %>% filter(released_year == 2022) %>%  group_by(released_month) %>%  
  summarise(num_of_song = n())
ggplot(data = month, mapping = aes(x = released_month, y = num_of_song)) + geom_col(fill = 'blue') +
  ggtitle('Number of songs per month in 2022') + labs(x = 'Month', y = 'Number of songs') + ylim(c(0, 85))
```

May had the highest number of Songs with 75 recorded. August had the least songs.

## What makes a Song a Hit?

We will examine the relationship between a song's stream and other factors 
#### Does music mode influence streams?

```{r}
plot(spotify$streams~spotify$mode, main = 'Streams by music mode', xlab = 'Music mode', ylab = 'Streams',
na.rm = TRUE)
```

Music mode doesn't seem to have an effect on streams.

#### Does danceability influence streams?

```{r}
ggplot(spotify, aes(x = danceability_., y = streams), na.rm = TRUE) + geom_point() + geom_smooth(method = 'lm') +
ggtitle('Relationship between streams and danceability') + labs(x = 'Danceability(%)', y = 'Streams')
cor1 <- cor(spotify$streams, spotify$danceability_., use = 'complete.obs') # checking correlation
cor1
```

There is a weak correlation between streams and danceability.

#### Do energy levels correlate with streams?

```{r}
ggplot(spotify, aes(x = energy_., y = streams)) + geom_point() + geom_smooth(method = 'lm') +
  ggtitle('Relationship between Streams and Energy Levels') + labs(x = 'Energy Level(%)', y = 'Streams')
cor2 <- cor(spotify$energy_., spotify$streams, use = 'complete.obs')
cor2
```

The correlation coefficient between the two variablesis negligible

#### Does instrumentalness correlate with streams?

```{r}
ggplot(spotify, aes(x = instrumentalness_., y = streams)) + geom_point() + geom_smooth(method = 'lm') + ggtitle('Relationship between Streams and Instrumentalness') + labs(x = 'Instrumentalness(%)', y = 'Streams')
cor3 <- cor(spotify$instrumentalness_., spotify$streams, use = 'complete.obs')
cor3
```

Negligible relationship as seen from the plot and correlation coefficient

#### Do Spoken words influence streams?

```{r}
ggplot(spotify, aes(x = speechiness_., y = streams)) + geom_point() + geom_smooth(method = 'lm') +
  ggtitle('Relationship between Streams and Spoken words') + labs(x = 'Spoken words(%)', y = 'Streams')
cor4 <- cor(spotify$speechiness_., spotify$streams, use = 'complete.obs')
cor4
```

Somewhat negative correlation.

### Multiple Linear Regression(measuring the effects of many variables)

```{r}
ml <- lm(streams~danceability_.+valence_.+energy_.+acousticness_.+instrumentalness_.+liveness_.+speechiness_., data = spotify)
summary(ml)
```

Most of the variables affecting streams in this linear regression model are statistically insignificant. The Low R-squared values show that the combination of these variables has a small effect on streams.

## Wrap up

Streams(popularity of an artist or song) are influenced by many quantitative and qualitative variables not accounted for in this dataset.
